Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-08 paper/paper.tex
pass over related work section

\section{Related Work}
\label{sec:related-work}
%
%\begin{itemize}
%  \item McBride devised a generalisation of Bove and Capretta's general
%recursive function monad~\cite{McBride15, BoveC01}
%  \item Guarded recursion for productive definitions and general recursion
%~\cite{AtkeyM13, PaviottiMB15}
%  \item PACO library~\cite{HurNDV13}
%  \item Sized Types \cite{HughesPS96, abs-1202-3496}
%  \item Copatterns \cite{AbelPTS13, AbelP16}
%  \item Not that related: extraction for mutable arrays~\cite{SAKAGUCHI2020102372}
%        \item Extraction of Distributed Mobile code in Coq~\cite{MiculanP12}.
%\end{itemize}
Encoding recursion schemes in Coq is not new. We compare our work with other
encodings of program calculation techniques in Coq (Section~\ref{sec:calccoq}),
recursion schemes in Coq (Sections~\ref{sec:calccoq} and~\ref{sec:reccoq}).
Finally, we compare our encoding of recursion schemes to other forms of
termination checking (Section~\ref{sec:termcheck}, and sized types
(Section~\ref{sec:sizedT}), as a way to guarantee termination of nonstructural
recursion.

@@

Within the domain of program optimization, program calculation serves as a
well-established programming technique aimed at deriving efficient programs
from their naive counterparts through systematic program
transformation~\cite{gibbonsSquiggol}. This area has been extensively explored
over the years. Tesson et al.\ demonstrated the efficacy of leveraging Coq to
establish an approach for implementing a robust system dedicated to verifying
the correctness of program transformations for functions that manipulate
lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and Emoto went further and
formalised recursion schemes in Coq~\cite{MurataE19}. Their development does
not include hylomorphisms and dynamorphisms, and relies on the functional
extensionality axiom, as well as further extensionality axioms for each
coinductive datatype that they use. They do not discuss the extracted OCaml
code from their formalisation.  Larchey-Wendling and Monin encode recursion
schemes in Coq, by formalising computational graphs of
algorithms~\cite{larchey2022braga}. Their work does not focus on encoding
generic recursion schemes, and proving their algebraic laws.  Castro-Perez et
al.~\cite{farmsCastro} encode the laws of hylomorphisms as part of the type
system of a functional language to calculate parallel programs from
specifications. Their work focuses on parallelism, and they do not formalise
their approach in a proof assistant, and the laws of hylomorphisms are axioms
in their system.

%\mvol{Need more general citations for program calculation.}

@@

@@

\subsubsection{Divide and Conquer Recursion}\label{sec:reccoq}

\subsubsection{Sized Types}\label{sec:sizedT}

Another approach to certifying termination of recursive functions is the use of
\emph{sized types}. Sized types were introduced by Hughes et al.\ as a way to
track/verify various properties of recursive programs, including productivity
and termination~\cite{HughesPS96}. The core idea of sized types is that types
express bounds on the sizes of recursive data structures. With this approach,
algorithms are implemented in the standard functional way, following the
divide-and-conquer pattern of splitting, recurring, and merging, with the
addition of the data types being indexed by a static approximation of the
relative \emph{size}. An advantage of this approach is that it allows for code
to be written in a natural way. There are costs, however. Programs must be
written in a way that accounts for the handling of size indexes, and support for
the size types must be added to the language.

This approach has been used to express nonstructurally recursive
algorithms in Agda: Copello et al.\ used sized types in a straightforward
formulation of mergesort~\cite{CopelloTB14}. Such an approach has also been used
in MiniAgda~\cite{abs-1202-3496}. A proposal exists to add sized types to
Coq as well, though it has not yet been adopted~\cite{ChanLB23}.

@@

Remaining axiom-free forces us to deal with the well-known
\emph{setoid hell}. As part of the future improvements, we will study how to
mitigate this problem. At the moment, we use a short ad-hoc tactic that is able
to automatically discharge many of these proofs in simple settings.  We will
study the more thorough and systematic use of proof automation for respectful
morphisms.  Generalised rewriting in proofs involving setoids tends to be quite
slow, due to the large size of the terms that need to be rewritten. Sometimes,
this size is hidden in implicit arguments and coercions. We will study
alternative formulations to try to improve the performance of the rewriting
tactics (e.g.  canonical structures).  Currently, Coq is unable to inline a
number of trivially inlineable definitions.  We will study alternative
definitions, or extensions to Coq's code extraction mechanisms to force the
full inlining of all container code that is used in hylomorphisms.  Finally,
proving termination still remains a hurdle. In our framework this reduces to
proving that the anamorphism
terminates in all inputs, and we provide a convenient connection to
well-founded recursion. Furthermore, recursive coalgebras compose with natural
transformations, which allows the reuse
of a number of core recursive coalgebras. A possible interesting future line of work
is the use of the approach by Abreu et al.~\cite{AbreuDHJMS23} in combination with ours to
improve code extraction from divide-and-conquer computations whose termination
does not require an external proof.

>>>>

\section{Related Work}
\label{sec:related-work}
%
%\begin{itemize}
%  \item McBride devised a generalisation of Bove and Capretta's general
%recursive function monad~\cite{McBride15, BoveC01}
%  \item Guarded recursion for productive definitions and general recursion
%~\cite{AtkeyM13, PaviottiMB15}
%  \item PACO library~\cite{HurNDV13}
%  \item Sized Types \cite{HughesPS96, abs-1202-3496}
%  \item Copatterns \cite{AbelPTS13, AbelP16}
%  \item Not that related: extraction for mutable arrays~\cite{SAKAGUCHI2020102372}
%        \item Extraction of Distributed Mobile code in Coq~\cite{MiculanP12}.
%\end{itemize}
Encoding recursion schemes in Coq is not new. We compare our work with other
encodings of program calculation techniques in Coq and recursion schemes in Coq
(Sections~\ref{sec:calccoq} and~\ref{sec:reccoq}). Additionally, we compare our
encoding of recursion schemes to other forms of termination checking
(Section~\ref{sec:termcheck}, as a way to guarantee termination of nonstructural
recursion.

@@

Within the domain of program optimization, program calculation serves as a
well-established programming technique aimed at deriving efficient programs from
their naive counterparts through systematic program
transformation~\cite{gibbonsSquiggol}. Programmers may specify algorithms
abstractly at a high level, and then derive an efficient implementation using
the process of program calculation~\cite{BirddeMoor96:Algebra}. This area has
been extensively explored over the years.

Tesson et al.\ demonstrated the efficacy of leveraging Coq to establish an
approach for implementing a robust system dedicated to verifying the correctness
of program transformations for functions that manipulate
lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and Emoto went further and
formalised recursion schemes in Coq~\cite{MurataE19}. Their development does not
include hylomorphisms and dynamorphisms, and relies on the functional
extensionality axiom, as well as further extensionality axioms for each
coinductive datatype that they use. They do not discuss the extracted OCaml code
from their formalisation. Larchey-Wendling and Monin encode recursion schemes in
Coq, by formalising computational graphs of algorithms~\cite{larchey2022braga}.
Their work does not focus on encoding generic recursion schemes, and proving
their algebraic laws. Castro-Perez et al.~\cite{farmsCastro} encode the laws of
hylomorphisms as part of the type system of a functional language to calculate
parallel programs from specifications. Their work focuses on parallelism, and
they do not formalise their approach in a proof assistant, and the laws of
hylomorphisms are axioms in their system.

%\mvol{Need more general citations for program calculation.}
\subsection{Divide and Conquer Recursion}\label{sec:reccoq}

@@

\mvol{Do we need to discuss this or can we cut it?}

@@

@@

Remaining axiom-free forces us to deal with the well-known
\emph{setoid hell}. As part of the future improvements, we will study how to
mitigate this problem. At the moment, we use a short ad-hoc tactic that is able
to automatically discharge many of these proofs in simple settings.  We will
study the more thorough and systematic use of proof automation for respectful
morphisms.  Generalised rewriting in proofs involving setoids tends to be quite
slow, due to the large size of the terms that need to be rewritten. Sometimes,
this size is hidden in implicit arguments and coercions. We will study
alternative formulations to try to improve the performance of the rewriting
tactics (e.g.\ canonical structures).  Currently, Coq is unable to inline a
number of trivially inlineable definitions.  We will study alternative
definitions, or extensions to Coq's code extraction mechanisms to force the
full inlining of all container code that is used in hylomorphisms.  Finally,
proving termination still remains a hurdle. In our framework this reduces to
proving that the anamorphism
terminates in all inputs, and we provide a convenient connection to
well-founded recursion. Furthermore, recursive coalgebras compose with natural
transformations, which allows the reuse
of a number of core recursive coalgebras. A possible interesting future line of work
is the use of the approach by Abreu et al.~\cite{AbreuDHJMS23} in combination with ours to
improve code extraction from divide-and-conquer computations whose termination
does not require an external proof.
